\textbf{Urban mobility in Latin America after COVID-19}

### 2024-01-27 manuscript/manuscript.tex
manuscript

\date{}
\begin{document}
\maketitle
\begin{abstract}
The COVID‐19 pandemic has impacted the national systems of population
movement around the world.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Spatial human mobility is key to creating sustainable, liveable and
inclusive cities. At the societal level, spatial mobility enables the
transfer of knowledge, skills and labour to places they are needed.
Spatial mobility also shapes service and transport demand across urban
spaces, and enables the monitoring and control of transmissible
diseases. At the individual level, mobility enables people to access and
achieve opportunities and aspirations in space. Understanding spatial
human mobility is thus important to supporting appropriate policy
responses to address societal challenges relating to carbon emissions,
urban planning, service delivery, public health, disaster management and
transport.

The COVID-19 pandemic resulted in a notable reduction in mobility within
cities. Nonpharmaceutical interventions to contain the spread of
COVID‐19 reduced overall levels of mobility. Especially during
lockdowns, mobility recorded reductions in the frequency, distance and
time of trips in cities across the globe. Rises in teleworking, online
schooling and remote shopping activity reduced the need to travel for
work, education, shopping and leisure. Coupled with fears of crowded
public spaces, nonpharmaceutical interventions prompted more
geographically localised mobility patterns.

@@

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook observed data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL) to define urban areas, and the Global Gridded Relative
Deprivation Index (GRDI) developed by NASA's Socioeconomic Data and
Applications Centre (SEDAC) from sociodemographic and satellite data
inputs. Building on existing evidence, we hypothesised that (1) urban
mobility has recovered returning to pre-pandemic levels of movement as
nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive business adopt hybrid
working.

@@

The evolution of the levels of movement was measured with respect to a
baseline period prior to the pandemic as described in
Section~\ref{sec-methods}. For the purposes of the analysis, we
aggregate the raw movement data temporally into months and spatially
into administrative units according to various GADM levels. The analysis
focuses on administrative areas that are within the boundaries of
functional urban areas as specified by the Global Human Settlement
Layer. For each administrative area, we compute the Relative Deprivation
Index based on data from NASA's Socioeconomic Data and Applications
Centre (SEDAC). Figure X displays the administrative areas included in
the study, coloured according to the Relative Index of Deprivation.
Predictions about the evolution of the levels of movement are made using
the Prophet forecasting procedure. Further details are provided in
Section~\ref{sec-methods}.

@@

We analysed the evolution of the relative intensity of urban mobility
with respect to a baseline period prior to the pandemic. Specifically,
movements covering a distance of at most 70 km are considered. For a
movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

Figure X displays the patterns of recovery for the mobility levels in
the administrative units belonging to functional urban areas in the
countries of interest. The three lines in each panel represent the mean
levels of mobility for administrative units grouped into one of three
terciles, according to their average relative deprivation index.

Generally, there was a drop in the levels of mobility with respect to
the baseline period in all four countries. This drop was especially
large for Argentina, Chile and Colombia, with Mexico displaying a
smaller decrease in the number of movements with respect to the
baseline. Following the initial drop in movement, all four countries
evolve towards the recovery of baseline levels of urban mobility, with a
generally increasing trend. There are fluctuations from the general
trend, which manifest differently for each country. These fluctuations
mirror each other in the case of Argentina and Colombia, where urban
mobility sharply bounces back closer to pre-pandemic levels around July
of 2020. Chile and Mexico display more progressive patterns of recovery,
although Chile never reaches baseline levels. These fluctuations are
unique to each country and can be attributed to local factors such as
the effects of seasonality or the different stringency measures imposed
by the national governments during the pandemic.

From Figure X, we observe that there is a consistent tendency in how
administrative units with varying levels of deprivation were affected by
the pandemic. For all four countries, we observe that the administrative
units in the most deprived tercile are the ones that experienced the
smallest loss in levels of mobility at the beginning of the pandemic.
Differences in the levels of mobility across relative deprivation
terciles diminish with time. Argentina and Chile stand out as the
countries with the largest differences in mobility levels for different
relative deprivation terciles.

@@

In this section we explore further the role of socioeconomic deprivation
in the evolution of the levels of urban mobility. For a given point in
time (i.e.~a month), we start by considering the relationship between
the number of movements relative to the pre-pandemic baseline period and
the average relative deprivation index, at the administrative unit
level. We assume that this relationship is linear and we use a linear
regression to estimate the slope and intercept characterising the line
of best fit. This is shown for April 2020 and March 2022 in the
right-hand side panels of Figure X. After obtaining the slope and
intercept for every month, we are able to plot the evolution of these
parameters for both the observed and forecasted data, as displayed on
the left-hand-side panels of the same Figure.

We find patterns in the evolution of the estimated parameters that
characterise the relationship between the levels of urban mobility and
RDI. In Argentina, Colombia and Mexico, we observe that the slope of
this relationship evolves to become smaller over time. The tendency is
not apparent in Chile, where the slope of the relationship remains
approximately the same despite the temporary fluctuations. The slope
captures the extent of differences in the level of urban mobility across
administrative units with varying levels of socioeconomic deprivation.
It can therefore be regarded as a measure of inequality. A slope equal
to zero would mean that all administrative units display the same
intensity of movement regardless of their socioeconomic deprivation
levels. Given the patterns observed in Argentina, Colombia and Mexico,
we find that at the beginning of the pandemic there were notable
inequalities between socioeconomic groups in terms of the levels of
urban mobility. While it has taken more than two years for Argentina and
Mexico to close the gap (their slope is close to zero from spring 2022),
inequalities persist in Chile and Colombia as of March 2023.

The intercept of the relationship displays stronger patterns, which are
consistent across the four countries. The intercept captures the urban
mobility levels in administrative where relative deprivation is zero, or
in other words, if there was no socioeconomic deprivation in an
administrative area, the intercept would represent its level of
mobility. The intercept was well below the baseline level at the early
stages of the pandemic. As observed in Figure X, while there are some
differences between countries in the evolution of the intercept, the
general tendency is for the intercept to increase. While Argentina and
Mexico reach values that are closer to the baseline towards the end of
the forecast period, the intercept for Chile and Colombia remains lower.
Therefore, if there were areas with no socioeconomic deprivation, we
would have seen a recovery in the levels of mobility, although not quite
back to baseline levels in the case of Chile and Colombia.

@@

>>>>

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities persist
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. Inequalities in mobility levels across
socioeconomic groups generally diminish over time, although they persist
in all the countries included in the analysis. This research highlights
the importance of understanding the long-term effects of the pandemic on
mobility and the need for equitable policy responses to address societal
challenges in urban areas.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Spatial human mobility is key to creating sustainable, livable and
inclusive cities. At the societal level, spatial mobility enables the
transfer of knowledge, skills and labour to places they are needed.
Spatial mobility also shapes service and transport demand across urban
spaces, and enables the monitoring and control of transmissible
diseases. At the individual level, mobility enables people to access and
achieve opportunities and aspirations in space. Understanding spatial
human mobility is thus important to supporting appropriate policy
responses to address societal challenges relating to carbon emissions,
urban planning, service delivery, public health, disaster management and
transport.

The COVID-19 pandemic resulted in a notable decrease in mobility,
particularly in cities. This decrease in the overall levels of mobility
was prompted by nonpharmaceutical interventions to contain the spread of
COVID‐19, coupled with general fears of visiting crowded public
spaces.Especially during lockdowns, mobility recorded reductions in the
frequency, distance and time of trips in cities across the globe. Higher
engagement with remote working, online schooling and shopping activity
reduced the need to travel for work, education, shopping and leisure,
hence giving rise to more geographically localised mobility patterns.

@@

Drawing on a dataset of 213 million observations from Meta-Facebook
users' mobile location data, we aim to assess socioeconomic differences
in the extent and persistence of decline in urban mobility in Argentina,
Chile, Colombia and Mexico during and after the COVID-19 pandemic from
March 2020 to March 2023. We use Meta-Facebook observed data to measure
origin-destination flows from March 2020 to May 2022, and Meta Prophet
time-series forecasting machine learning algorithm to predict
origin-destination flows from June 2022 to March 2023. We use Functional
Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
(GHSL) to define urban areas, and the Global Gridded Relative
Deprivation Index (GRDI) developed by NASA's Socioeconomic Data and
Applications Centre (SEDAC) from sociodemographic and satellite data
inputs. Building on existing evidence, we hypothesised that (1) urban
mobility has recovered returning to pre-pandemic levels of movement as
nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
differences in urban mobility have endured the pandemic reflecting deep
societal inequalities as knowledge-intensive businesses adopt hybrid
working.

@@

The evolution of the levels of movement is measured with respect to a
baseline period prior to the pandemic as described in
Section~\ref{sec-methods}. For the purposes of the analysis, we
aggregate the raw movement data temporally into months and spatially
into administrative units according to various GADM levels. The analysis
focuses on administrative areas that are within the boundaries of
functional urban areas as specified by the Global Human Settlement
Layer. For each administrative area, we compute the Relative Deprivation
Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative areas
included in the study, coloured according to the Relative Index of
Deprivation. Predictions about the evolution of the levels of movement
are made using the Prophet forecasting procedure. Further details are
provided in Section~\ref{sec-methods}.

\begin{figure}

{\centering \includegraphics{figures/maps.pdf}

}

\caption{Administrative units within functional urban areas in
Argentina, Chile, Colombia and Mexico, coloured according to the average
Relative Deprivation Index.}

\end{figure}

@@

We analyse the evolution of the relative intensity of urban mobility
with respect to a baseline period prior to the pandemic. Specifically,
movements covering a distance of at most 70 km are considered. For a
movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

Figure 2 displays the patterns of recovery for the mobility levels in
the administrative units belonging to functional urban areas in the
countries of interest. The three lines in each panel represent the mean
levels of mobility for administrative units grouped into one of three
terciles, according to their average relative deprivation index.

\begin{figure}

{\centering \includegraphics{figures/prediction-rdi-band.pdf}

}

\caption{Patterns of recovery for urban mobility in administrative units
from Argentina, Chile, Colombia and Mexico.}

\end{figure}

Generally, there was a drop in the levels of mobility with respect to
the baseline period in all four countries. This drop was especially
large for Argentina, Chile and Colombia, with Mexico displaying a
smaller decrease in the relative number of movements with respect to the
baseline. Following the initial drop in movement, all four countries
evolve towards the recovery of baseline levels of urban mobility, with a
generally increasing trend. There are, however, fluctuations from the
general trend which manifest differently for each country. These
fluctuations mirror each other in the case of Argentina and Colombia,
where urban mobility sharply bounces back closer to pre-pandemic levels
around July of 2020. Chile and Mexico display more progressive patterns
of recovery, although Chile never reaches baseline levels. These
fluctuations are unique to each country and can be attributed to local
factors such as the effects of seasonality or the different stringency
measures imposed by the national governments during the pandemic.

From Figure 2, we observe that there is a consistent tendency in how
administrative units with varying levels of deprivation were affected by
the pandemic. For all four countries, we observe that the administrative
units in the most deprived tercile are the ones that experienced the
smallest loss in levels of mobility at the beginning of the pandemic.
Differences in the levels of mobility across relative deprivation
terciles diminish with time. Argentina and Chile stand out as the
countries with the largest differences in mobility levels for different
relative deprivation terciles.

@@

In this section we explore further the role of socioeconomic deprivation
in the evolution of the levels of urban mobility. For a given point in
time (i.e.~a month), we start by considering the relationship between
the number of movements relative to the pre-pandemic baseline period and
the average relative deprivation index, at the administrative unit
level. We assume that this relationship is linear and we use a linear
regression to estimate the slope and intercept characterising the line
of best fit. This is shown for April 2020 and March 2022 in the
right-hand side panels of Figure 3. After obtaining the slope and
intercept for every month, we are able to plot the evolution of these
parameters for both the observed and forecasted data, as displayed on
the left-hand-side panels of the same Figure.

\begin{figure}

{\centering \includegraphics{figures/regression-evo-nobackground.pdf}

}

\caption{Evolution of the parameters characterising the relationship
between relative deprivation index and number of movements relative to
the pre-pandemic baseline period.}

\end{figure}

We find patterns in the evolution of the estimated parameters that
characterise the relationship between the levels of urban mobility and
RDI. In Argentina, Colombia and Mexico, we observe that the slope of
this relationship evolves to become smaller over time. The tendency is
not apparent in Chile, where the slope of the relationship remains
approximately the same despite the temporary fluctuations. The slope
captures the extent of differences in the level of urban mobility across
administrative units with varying levels of socioeconomic deprivation.
It can therefore be regarded as a measure of inequality in mobility
patterns across socioeconomic groups. A slope equal to zero would mean
that all administrative units display the same intensity of movement
regardless of their socioeconomic deprivation levels. Given the patterns
observed in Argentina, Colombia and Mexico, we find that at the
beginning of the pandemic there were notable inequalities between
socioeconomic groups in terms of the levels of urban mobility. While it
has taken more than two years for Argentina and Mexico to close the gap
(their slope is close to zero from spring 2022), inequalities persist in
Chile and Colombia as of March 2023.

The intercept of the relationship displays stronger patterns, which are
consistent across the four countries. The intercept estimates the urban
mobility levels that would be observed in administrative units where
relative deprivation is zero. The intercept was below the baseline level
at the early stages of the pandemic. As observed in Figure 3, while
there are some differences between countries in the evolution of the
intercept, the general tendency is for the intercept to increase. While
Argentina and Mexico reach values that are closer to the baseline
towards the end of the forecast period, the intercept for Chile and
Colombia remains lower. Therefore, if there were areas with no
socioeconomic deprivation, we would have seen a recovery in the levels
of mobility, especially in Argentina and Mexico

@@

Using location data from Meta-Facebook users, our study aimed to examine
the evolution of patterns of mobility across socioeconomic groups in
functional urban areas from Argentina, Chile, Colombia and Mexico from
April 2020 to March 2023, following the COVID-19 pandemic. We found a
systematic decline in population movement levels during the
implementation of non-pharmaceutical interventions in April 2020, with
the largest reductions observed in the most affluent administrative
units within functional urban areas from Argentina, Chile and Colombia.
While mobility rebounded closer to pre-pandemic levels approximately two
years later after COVID-19 restrictions eased, movement intensity
remained below pre-pandemic levels in Chile and generally in areas where
the relative deprivation index is lower. Furthermore, we found that at
the beginning of the pandemic there were inequalities between
socioeconomic groups in terms of the levels of urban mobility. While it
has taken more than two years for Argentina and Mexico to gradually
close the gap, inequalities persist in Chile and Colombia as of March
2023 according to our estimated data.

We focused the analysis on short-distance movements in urban areas,
which are typically considered to capture local and routine mobility
(Owen and Green 1992). However, due to the characteristics of the
dataset, we are unable to distinguish movements of this nature from
short-distance movements that involve a permanent change of place of
residence. Future work could extend and validate our work identifying
changes to moves relating permanent and temporary displacements, and
triangulating alternative DFD and traditional data sources, such as
censuses and representative surveys collecting information on
residential location. Further research is also needed to establish the
cause of the observed changes in the spatial patterns of internal
population movement since the start of the COVID-19 pandemic.
Understanding these causes can assist in anticipating long-term
structural changes in the direction of mobility flows in the coming
years.

Additionally, accessing and working with DFD to measure human mobility
and migration is challenging. DFD are known to suffer from biases and
issues of statistical representation reflecting differences in
technology usage and accessibility (Rowe 2023). While concerted efforts
to develop data services and methodological frameworks to leverage on
DFD exist in the Global North {[}REF{]}, limited progress has been done
in Latin America. National government and statistical agencies have a
prime position to promote the creation of data sharing partnerships with
private companies and create the digital infrastructure to ensure equal
and secure access to DFD sources for the social good. With academia,
these agencies could play a major role in promoting the creation of a
methodological framework for the use and analysis of DFD, setting common
definitions, metrics, and standard approaches to measure and correct
data biases and validate data analyses. Location data have been proven
key to monitor and develop appropriate policy responses to population
displacement and movement during natural disasters, wars and epidemics
(Green, Pollock, and Rowe 2021). We envision resources like those
created by the European Commission Joint Research Centre (2022) and the
UNSD (2019), identifying sources of non-traditional data and
establishing methodological guidelines for the use of mobile phone data
for official mobility statistics. Existing resources have a global
character, we call for initiatives with a regional focus accounting for
differences in the local landscape of data availability and digital
technology usage.
